Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support).

### 2017-02-10 csf-2017/paper.tex
added lemma for f2
Some work on explaining the stack and return pointer management discipline.

% Instructions
% No arguments
\newcommand{\zinstr}[1]{\mathtt{#1}}
\newcommand{\fail}{\zinstr{fail}}
\newcommand{\halt}{\zinstr{halt}}
% One argument
\newcommand{\oneinstr}[2]{\zinstr{#1} \; #2}
\newcommand{\jmp}[1]{\oneinstr{jmp}{#1}}
% Two arguments
\newcommand{\twoinstr}[3]{\zinstr{#1} \; #2 \; #3}
\newcommand{\restricttwo}[2]{\twoinstr{restrict}{#1}{#2}}
\newcommand{\jnz}[2]{\twoinstr{jnz}{#1}{#2}}
\newcommand{\isptr}[2]{\twoinstr{isptr}{#1}{#2}}
\newcommand{\geta}[2]{\twoinstr{geta}{#1}{#2}}
\newcommand{\getb}[2]{\twoinstr{getb}{#1}{#2}}
\newcommand{\gete}[2]{\twoinstr{gete}{#1}{#2}}
\newcommand{\getp}[2]{\twoinstr{getp}{#1}{#2}}
\newcommand{\getl}[2]{\twoinstr{getl}{#1}{#2}}
\newcommand{\move}[2]{\twoinstr{move}{#1}{#2}}
\newcommand{\store}[2]{\twoinstr{store}{#1}{#2}}
\newcommand{\load}[2]{\twoinstr{load}{#1}{#2}}
\newcommand{\lea}[2]{\twoinstr{lea}{#1}{#2}}
% Three arguments
\newcommand{\threeinstr}[4]{\zinstr{#1} \; #2 \; #3 \; #4}
\newcommand{\restrict}[3]{\threeinstr{restrict}{#1}{#2}{#3}}
\newcommand{\subseg}[3]{\threeinstr{subseg}{#1}{#2}{#3}}
\newcommand{\plus}[3]{\threeinstr{plus}{#1}{#2}{#3}}

@@

The instruction set of our machine is rather basic. Instructions $i$ include
relatively standard jump ($\jmp{}$), conditional jump ($\jnz{}{}$) and move
($\move{}{}$, copies words between registers) instructions. Also familiar are
load and store instructions for reading from and writing to memory ($\load{}{}$
and $\store{}{}$) and an arithmetic addition operator ($\plus{}{}{}$). There are
three instructions for modifying capabilities: $\lea{}{}$ (modifies the current
address), $\restrict{}{}$ (modifies the permission and local/global flag) and
$\subseg{}{}{}$ (modifies the range of a capability). Importantly, these
instructions take care that the resulting capability always carries less
authority than the original (e.g. restrict will only weaken a permission).
Finally, there is an instruction $\isptr{}{}$ for testing whether a word is a
capability or a number and instructions $\getp{}{}$, $\getl{}{}$, $\getb{}{}$,
$\gete{}{}$ and $\geta{}{}$ for obtaining respectively the permissions,
local/global flag, base, end and current address of a capability.

@@

The $\load{}{}$ instruction loads the contents of the requested memory location
into the appropriate register, but only if the capability has appropriate
authority (i.e. read permission and an appropriate range). The $\restrict{}{}$
instruction will update a capability's permissions and global/local status in
the register file, but requires that the new permissions are weaker than the
original and does not turn local capabilities into global ones. The $\geta{}{}$
instruction queries the current address of a capability and stores it in another
register.

@@

@@

The basic idea of our approach is actually rather simple: we stick to a single,
rather standard C stack and register-passed return pointers in much the same way
as a standard C ABI. However, to prevent various ways of misusing this basic
scheme, we put local capabilities to work and need to perform a number of
not-always-obvious checks and safety measures.

\paragraph{Invoking the adversary}
* stack and return pointer as local caps
* clearing registers
* clearing the stack

\paragraph{Returning from the adversary}
* return pointer as enter pointer to activation record on the stack.
* activation record contains previous pc and stack pointer

>>>>

% Instructions
% No arguments
\newcommand{\zinstr}[1]{\mathtt{#1}}
\newcommand{\fail}{\zinstr{fail}}
\newcommand{\halt}{\zinstr{halt}}
% One argument
\newcommand{\oneinstr}[2]{
  \ifthenelse{\equal{#2}{}}
  {\zinstr{#1}}
  {\zinstr{#1} \; #2}
}
\newcommand{\jmp}[1]{\oneinstr{jmp}{#1}}
% Two arguments
\newcommand{\twoinstr}[3]{
  \ifthenelse{\equal{#2#3}{}}
  {\zinstr{#1}}
  {\zinstr{#1} \; #2 \; #3}
}
\newcommand{\restricttwo}[2]{\twoinstr{restrict}{#1}{#2}}
\newcommand{\jnz}[2]{\twoinstr{jnz}{#1}{#2}}
\newcommand{\isptr}[2]{\twoinstr{isptr}{#1}{#2}}
\newcommand{\geta}[2]{\twoinstr{geta}{#1}{#2}}
\newcommand{\getb}[2]{\twoinstr{getb}{#1}{#2}}
\newcommand{\gete}[2]{\twoinstr{gete}{#1}{#2}}
\newcommand{\getp}[2]{\twoinstr{getp}{#1}{#2}}
\newcommand{\getl}[2]{\twoinstr{getl}{#1}{#2}}
\newcommand{\move}[2]{\twoinstr{move}{#1}{#2}}
\newcommand{\store}[2]{\twoinstr{store}{#1}{#2}}
\newcommand{\load}[2]{\twoinstr{load}{#1}{#2}}
\newcommand{\lea}[2]{\twoinstr{lea}{#1}{#2}}
% Three arguments
\newcommand{\threeinstr}[4]{
  \ifthenelse{\equal{#2#3#4}{}}
  {\zinstr{#1}}
  {\zinstr{#1} \; #2 \; #3 \; #4}
}
\newcommand{\restrict}[3]{\threeinstr{restrict}{#1}{#2}{#3}}
\newcommand{\subseg}[3]{\threeinstr{subseg}{#1}{#2}{#3}}
\newcommand{\plus}[3]{\threeinstr{plus}{#1}{#2}{#3}}

@@

The instruction set of our machine is rather basic. Instructions $i$ include
relatively standard jump ($\jmp{}$), conditional jump ($\jnz{}{}$) and move
($\move{}{}$, copies words between registers) instructions. Also familiar are
load and store instructions for reading from and writing to memory ($\load{}{}$
and $\store{}{}$) and an arithmetic addition operator ($\plus{}{}{}$). There are
three instructions for modifying capabilities: $\lea{}{}$ (modifies the current
address), $\restricttwo{}{}$ (modifies the permission and local/global flag) and
$\subseg{}{}{}$ (modifies the range of a capability). Importantly, these
instructions take care that the resulting capability always carries less
authority than the original (e.g. restrict will only weaken a permission).
Finally, there is an instruction $\isptr{}{}$ for testing whether a word is a
capability or a number and instructions $\getp{}{}$, $\getl{}{}$, $\getb{}{}$,
$\gete{}{}$ and $\geta{}{}$ for obtaining respectively the permissions,
local/global flag, base, end and current address of a capability.

@@

The $\load{}{}$ instruction loads the contents of the requested memory location
into the appropriate register, but only if the capability has appropriate
authority (i.e. read permission and an appropriate range). The $\restricttwo{}{}$
instruction will update a capability's permissions and global/local status in
the register file, but requires that the new permissions are weaker than the
original and does not turn local capabilities into global ones. The $\geta{}{}$
instruction queries the current address of a capability and stores it in another
register.

@@

Explain malloc
\dominique{not sure what we want to say about it here either...}

@@

The basic idea of our approach is quite simple: we stick to a single, rather
standard C stack and register-passed return pointers in much the same way as a
standard C ABI. However, to prevent various ways of misusing this basic scheme,
we put local capabilities to work and take a number of not-always-obvious safety
measures. In the next paragraphs, we will explain the issues to be considered in
all the relevant situations: when (1) starting our program, (2) invoking the
adversary, (3) returning from the adversary, (4) invoking an adversary callback
and (5) having a callback invoked by the adversary.

\dominique{Add a summary table somewhere?}

\paragraph{Program startup}
When our code starts executing, we assume that the language runtime initializes
the memory as follows: a contiguous array of memory is reserved for the stack,
for which we receive a stack pointer in a register that we reserve for this
purpose \dominique{give this register a special name like sp?}. This stack
pointer must be local and it must carry $\rwlx$ permission. It is crucial for
our approach that the stack is the only part of memory for which there will ever
be capabilities with $\rwlx$ or $\rwl$ capabilities. We expect the language
runtime to enforce this by not handing out such capabilities for other parts of
memory in any situation (i.e. not through malloc, not during loading or linking
etc.). Additionally, in our examples, we typically also assume some static parts
of memory where instructions are stored or static data. Another part of memory
(called the heap) is initially governed by malloc and at program startup, no
other code has capabilities for this memory. Malloc will hand out capabilities
for allocated regions as requested, but it will never hand out capabilities with
write-local capability (i.e. no $\rwlx$ or $\rwl$ permissions). For simplicity,
we assume that memory allocated through malloc cannot be freed.

\paragraph{Invoking the adversary}
During execution, we use the stack as normal: we may push values onto it and pop
them back off (e.g. for register spilling etc.). When we invoke the adversary,
we need to take great care to only provide him with the authority we want him to
have. Specifically, we clear all the registers, except those that contain
arguments the adversary is intended to have access to. We leave a copy of the
stack pointer in the appropriate register, but only after we have used the
$\subseg{}{}{}$ to shrink its authority to just the part of the stack that we
are not using. In one of the registers, we will also provide a return pointer.
We will go into constructing the return pointer further below.

It is already important to note that the return pointer
will be provided to the adversary as local capabilities. If the adversary were
able to store away the return pointer in a global data structure, then he would
be able to jump to it in circumstances where this should not be possible: he
may, for example, store the return pointer, legally jump to it a first time,
wait to be invoked again and then jump to the old return pointer a second time,
rather than the new return pointer received for the new invocation. Similarly,
he may store the return pointer, invoke a function in the trusted code, wait to
be invoked again and then jump to the stored return pointer rather than the new
return pointer for the second invocation.

Because the return pointer is local, the adversary is only able to store it
through write-local capabilities, which (because of our assumptions above)
means: on the stack. However, this makes it crucial that the stack pointer is
also local: if the adversary were able to store the stack pointer in a global
data structure, he/she can store return pointers on the stack, and play similar
mischief as mentioned above. Storing away stack pointers for use during future
invocations would also be dangerous for other reasons: during such a future
invocation, a larger part of the stack may be in use by the trusted code, which
the adversary may still have access to through the old stack pointer.

A final measure we need to take is that before invocation of the adversary, we
need to clear (i.e. overwrite with zeros) the part of the stack that we are not
using.  Otherwise, that part of the stack may still contain data and
capabilities from previously executed code, including local capabilities since
they are allowed on the stack.  The adversary would be able to take advantage of
this by using it to store either stack pointers or return pointers from previous
invocations and play the same tricks as above.

\paragraph{Returning from the adversary}
So return pointers must be passed as local capabilities. But what should their
permissions be, what memory should they point to and what should that memory
(let's call it the activation record) contain? Let us answer the last question
first by considering what should happen when the adversary jumps to a return
pointer and what capabilities we need for achieving that. Essentially, the
program counter should be restored to the code after the call into the
adversary, so the activation record should store this program counter.
Additionally, the stack pointer should also be restored to its original value.
Since the adversary has a more restricted authority over the stack than the code
making the call, we cannot hope to reconstruct the return pointer from the stack
pointer owned by the adversary. Instead, the old stack pointer should also be
stored as part of the activation record.

\begin{figure}
  \begin{tikzpicture}
    \draw[fill=gray!20,draw=none] (0,4) rectangle (4,3.5);
    \draw[fill=gray!20,draw=none] (0,0.5) rectangle (4,1);
    \draw (0,0.5) -- (0,4);
    \draw (4,0.5) -- (4,4);
    \draw (0,3.5) rectangle (4,3) node[pos=.5] {\footnotesize previous program counter};
    \draw (0,3) rectangle (4,2.5) node[pos=.5] {\footnotesize previous stack pointer};
    \draw (0,2.5) rectangle (4,1) node[pos=.5] {\footnotesize restore instructions};
    \draw (-2,2.25) node (rp) {return pointer};
    \draw[->] (rp.east) -- (-.2,2.25);
  \end{tikzpicture}

  \caption{Structure of an activation record}
  \label{fig:activ-rec-struct}
\end{figure}

It's important that neither of these capabilities are accessible to the
adversary. In other words, we need to provide the adversary a capability that he
can jump to but not read from, i.e. an enter capability. To make this work, we
construct the activation record as depicted in Figure~\ref{fig:activ-rec-struct}.

* return pointer as enter pointer to activation
record on the stack. * activation record contains previous pc and stack pointer
